Estimating SARS-CoV-2 transmission in England from randomly sampled prevalence surveys

### 2024-03-27 MCMC-appendix.tex
Fix refs
finish proof reading

MCMC (Markov chain Monte Carlo) is a class of algorithms for sampling from a distribution where the normalising constant for the distribution is unknown.
I use it to sample a set of parameters, say $\psi$ (which may be a scalar or vector), from a posterior distribution $p(\psi \mid y)$ where $y$ is the data (again, scalar or vector).
MCMC algorithms produce a sample $\psi^1, \psi^2, \dots, \psi^M$ from $p(\psi \mid y)$ by constructing a Markov chain with $p(\psi \mid y)$ as its stationary distribution~\autocite[275]{gelmanBDA}.
It can be proved that as $M\to\infty$, the sample tends towards a sample from the stationary distribution, \ie the target posterior.
This appendix outlines the basic theory and implementation of Markov chain Monte Carlo (MCMC) methods.
Any reader interested in more technical details or proofs of the results here should consult the references within.

@@

The proposal distribution $q$ is most commonly a (possibly multivariate) normal distribution centred on $\psi$; this scheme is used in \cref{E-SEIR}.
Tuning the proposal distribution to propose ``good'' moves is a key part of MCMC.
The ideal proposal distribution is the posterior of interest, however, this is impossible because that is what MCMC is trying to estimate~\autocite[296]{gelmanBDA}.
One alternative is \emph{adaptive} algorithms which tune the proposal distribution based on previous draws in an attempt to approximate the posterior; I use such a scheme in \cref{E-SEIR:sec:inference}.

@@

\subsection{Stability}

>>>>

MCMC (Markov chain Monte Carlo) is a class of algorithms which can sample from a distribution where the normalising constant for the distribution is unknown.
I use it to sample a set of parameters, say $\psi$ (which may be a scalar or vector), from a posterior distribution $p(\psi \mid y)$ where $y$ is the data (again, scalar or vector).
MCMC algorithms produce a sample $\psi^1, \psi^2, \dots, \psi^M$ from $p(\psi \mid y)$ by constructing a Markov chain with $p(\psi \mid y)$ as its stationary distribution~\autocite[275]{gelmanBDA}.
It can be proved that as $M\to\infty$, the sample tends towards a sample from the stationary distribution, \ie the target posterior.
This appendix outlines the basic theory and implementation of Markov chain Monte Carlo (MCMC) methods.
Any reader interested in more technical details or proofs of the results here should consult the references within.

@@

The proposal distribution $q$ is often a (possibly multivariate) normal distribution centred on $\psi$; this scheme is used in \cref{E-SEIR}.
Tuning the proposal distribution to propose ``good'' moves is a key part of MCMC.
The ideal proposal distribution is the posterior of interest, however, this is impossible because that is what MCMC is trying to estimate~\autocite[296]{gelmanBDA}.
One alternative is \emph{adaptive} algorithms which tune the proposal distribution based on previous draws in an attempt to approximate the posterior; I use such a scheme in \cref{E-SEIR:sec:inference}.

@@

\subsection{Stability} \label{MCMC:sec:stability}

### 2024-03-27 cis-perfect-testing.tex
Overleaf: mostly ch 5 corrections
proof ch 6 inc new figures
Updates from Overleaf
finish proof reading

By definition (see \cpageref{perf-test:def:admissible}), we have, for any $i \in \set{D}$:
\begin{align}
p_{ia} =& \prob((B_j, E_j) \in \alpha_i \mid \vec{\theta}, i(j) = i) \\
\alpha_i =& \{ (b, e) : l_{j(i)}^{(b)} \leq b \leq r_{j(i)}^{(b)} \wedge l_{j(i)}^{(e)} \leq e \leq r_{j(i)}^{(e)}\}.
\intertext{Omitting the conditioning on $\vec{\theta}$ and $i(j) = i$, this gives:}
p_{ia}
=& \prob \left( l_{j(i)}^{(b)} \leq B_{j} \leq r_{j(i)}^{(b)}, l_{j(i)}^{(e)} \leq E_{j} \leq r_{j(i)}^{(e)} \right) \\
=& \prob \left( l_{j(i)}^{(e)} \leq E_{j} \leq r_{j(i)}^{(e)} \mid l_{j(i)}^{(b)} \leq B_{j} \leq r_{j(i)}^{(b)} \right) \\
   &\times\prob \left( l_{j(i)}^{(b)} \leq B_{j} \leq r_{j(i)}^{(b)} \right) \\
=& \sum_{b = l_{j(i)}^{(l)}}^{r_{j(i)}^{(b)}} \prob \left( l_{j(i)}^{(e)} \leq E_{j} \leq r_{j(i)}^{(e)} \mid B_{j} = b \right) \prob \left(B_{j} = b \right) \\
=& \sum_{b = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \prob \left( l_{j(i)}^{(e)} - b + 1 \leq D_{j} \leq r_{j(i)}^{(e)} - b + 1 \right) \prob \left(B_{j} = b \right) &\text{by def of $D_{j}$} \\
=& \sum_{b = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \left( S_{\vec{\theta}}(l_{j(i)}^{(e)} - b + 1) - S_{\vec{\theta}}(r_{j(i)}^{(e)} - b + 2) \right) \prob \left(B_{j} = b \right) &\text{by def of $S_{\vec{\theta}}$} \\
\propto& \sum_{b = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \left( S_{\vec{\theta}}(l_{j(i)}^{(e)} - b + 1) - S_{\vec{\theta}}(r_{j(i)}^{(e)} - b + 2) \right)
\end{align}
under the assumption of uniform probability of infection time.
This is the standard form of the likelihood for doubly interval censored data without truncation~\autocite[e.g.][]{sunEmpirical}.

@@

% An episode is undetected if and only if no tests are performed during the episode or if there was no negative test prior to the episode.
% Equivalently, that the first test at or after $b$ is after $e$, or that there is no negative test prior to $b$.
To formalise this, first define $\tau_{\sched_i}(t)$ as the time until the next test at or after time $t$ in the schedule $\sched_i$:
\begin{align}
\tau_{\sched_i}(t) &= \min \{ t' \in \sched_i : t' \geq t \} - t
\end{align}
defining $\min \emptyset = \infty$; \ie $\tau_{\sched_i}(t) = \infty$ if there is no $t' \in \sched_i$ such that $t' \geq t$.
The first condition can now be expressed as $e_j \geq b_j + \tau_{\sched_{i(j)}}(b_j)$.
Equivalently, $d_j \geq \tau_{\sched_{i(j)}}(b_j) + 1$.

% Then $\Omega_i$ can be written as:
% \begin{align}
% \Omega_i = \{ (b, e) \ssep \tau_{\sched_i}(b) + b > e \vee b \leq \min(\sched_i) \}.
% \end{align}
Therefore, omitting the conditioning on $\vec{\theta}$ and $i(j) = i$:
\begin{align}
1 - p_{iu}
&= 1 - \prob((B_{j}, E_{j}) \in \Omega_i) \\
&= \prob(D_j \geq \tau_{\sched_{i}}(B_j)+ 1, B_j > \min \sched_{j}) \\
&= \sum_{b = \min \sched_{i} + 1}^T \prob(D_j \geq \tau_{\sched_{i}}(b) + 1 \mid B_j = b) \prob(B_j = b)\\
&= \frac{1}{T} \sum_{b = \min \sched_{i} + 1}^T S_{\vec{\theta}}(\tau_{\sched_{i}}(b) + 1)
\label{perf-test:eq:piu}
\end{align}
defining $S_{\vec{\theta}}(\infty) = 0$.

@@

I implemented the simulation in R~4.2.0~\autocite{R-4-2-0} using tidyverse~2.0.0~\autocite{tidyverse} and targets~1.1.3~\autocite{targetsPackage}.

>>>>

By definition (see \cpageref{perf-test:def:admissible}), we have, for any $i \in \set{D}$:
\begin{align}
p_{ia} =& \prob((B_j, E_j) \in \alpha_i \mid \vec{\theta}, i(j) = i) \\
\alpha_i =& \{ (b, e) : l_{j(i)}^{(b)} \leq b \leq r_{j(i)}^{(b)} \wedge l_{j(i)}^{(e)} \leq e \leq r_{j(i)}^{(e)}\}.
\intertext{Omitting the conditioning on $\vec{\theta}$ and $i(j) = i$, this gives:}
p_{ia}
=& \prob \left( l_{j(i)}^{(b)} \leq B_{j} \leq r_{j(i)}^{(b)}, l_{j(i)}^{(e)} \leq E_{j} \leq r_{j(i)}^{(e)} \right) \\
=& \prob \left( l_{j(i)}^{(e)} \leq E_{j} \leq r_{j(i)}^{(e)} \mid l_{j(i)}^{(b)} \leq B_{j} \leq r_{j(i)}^{(b)} \right) \\
   &\times\prob \left( l_{j(i)}^{(b)} \leq B_{j} \leq r_{j(i)}^{(b)} \right) \\
=& \sum_{b = l_{j(i)}^{(l)}}^{r_{j(i)}^{(b)}} \prob \left( l_{j(i)}^{(e)} \leq E_{j} \leq r_{j(i)}^{(e)} \mid B_{j} = b \right) \prob \left(B_{j} = b \right) \\
=& \sum_{b = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \prob \left( l_{j(i)}^{(e)} - b + 1 \leq D_{j} \leq r_{j(i)}^{(e)} - b + 1 \right) \prob \left(B_{j} = b \right) &\text{by def of $D_{j}$} \\
=& \sum_{b = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \left( S_{\vec{\theta}}(l_{j(i)}^{(e)} - b + 1) - S_{\vec{\theta}}(r_{j(i)}^{(e)} - b + 2) \right) \prob \left(B_{j} = b \right) &\text{by def of $S_{\vec{\theta}}$} \\
\propto& \sum_{b = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \left( S_{\vec{\theta}}(l_{j(i)}^{(e)} - b + 1) - S_{\vec{\theta}}(r_{j(i)}^{(e)} - b + 2) \right)
\label{perf-test:eq:pia}
\end{align}
under the assumption of uniform probability of infection time.
This is the standard form of the likelihood for doubly interval censored data without truncation~\autocite[e.g.][]{sunEmpirical}.

@@

% An episode is undetected if and only if no tests are performed during the episode or if there was no negative test prior to the episode.
% Equivalently, that the first test at or after $b$ is after $e$, or that there is no negative test prior to $b$.
To formalise this, first define $\tau_{\sched_i}(t)$ as the time until the next test at or after time $t$ in the schedule $\sched_i$:
\begin{align}
\tau_{\sched_i}(t) &= \min \{ t' \in \sched_i : t' \geq t \} - t
\label{perf-test:eq:tau-def}
\end{align}
defining $\min \emptyset = \infty$; \ie $\tau_{\sched_i}(t) = \infty$ if there is no $t' \in \sched_i$ such that $t' \geq t$.
The first condition can now be expressed as $e_j \geq b_j + \tau_{\sched_{i(j)}}(b_j)$.
Equivalently, $d_j \geq \tau_{\sched_{i(j)}}(b_j) + 1$.
The third condition can be expressed as $\tau_{\sched_{i(j)}}(b_j) < \infty$.

% Then $\Omega_i$ can be written as:
% \begin{align}
% \Omega_i = \{ (b, e) \ssep \tau_{\sched_i}(b) + b > e \vee b \leq \min(\sched_i) \}.
% \end{align}
Therefore, omitting the conditioning on $\vec{\theta}$ and $i(j) = i$:
\begin{align}
1 - p_{iu}
&= 1 - \prob((B_{j}, E_{j}) \in \Omega_i) \\
&= \prob(D_j \geq \tau_{\sched_{i}}(B_j)+ 1, \tau_{\sched_{i(j)}}(b_j) < \infty, B_j > \min \sched_{j}) \\
&= \sum_{b = \min \sched_{i} + 1}^T \prob(D_j \geq \tau_{\sched_{i}}(b) + 1, \tau_{\sched_{i(j)}}(b_j) < \infty \mid B_j = b) \prob(B_j = b)\\
&= \frac{1}{T} \sum_{b = \min \sched_{i} + 1}^T S_{\vec{\theta}}(\tau_{\sched_{i}}(b) + 1)
\label{perf-test:eq:piu}
\end{align}
defining $S_{\vec{\theta}}(\infty) = 0$.

@@

I implemented the simulation in R~4.2.0~\autocite{R-4-2-0} using tidyverse~2.0.0~\autocite{tidyverse}.

### 2024-03-27 incidence-prevalence.tex
finish proof reading

Considering multiple values of $t$ makes \cref{inc-prev:eq:EPt} into a system of simultaneous equations.
Therefore, the system can be written in matrix form:
\begin{align}
  \E(\vec{P} \mid \vec{Z}) = \matr{S} \vec{Z}. \label{inc-prev:eq:EPt-matrix} \\
\intertext{$\matr{S}$ is a lower-triangular matrix formed from $S$:}
    (\matr{S})_{i,j} &= \begin{cases}
        S(i - j + 1) & i - \dmax - 1 \leq j \leq i \\
        0 & \text{otherwise.}
    \end{cases} \label{inc-prev:eq:Smatrix}
\end{align}
In \cref{E-backcalc}, I will need to solve for this system for $\vec{Z}$ given a known $\hat{P} = \E(\vec{P} \mid {Z})$.
As $\matr{S}$ is lower-triangular, I use forward-substitution~\autocite{cormenMatrix}.
Forward-substitution sequentially solves the $t$th element of $\vec{Z}$ ($t = 1, \dots, T$) by calculating:
\begin{align}
Z_t
&= \frac{\hat{P}_t - \sum_{i=1}^{t-1} S_{ti} Z_i}{S_{tt}} \\
&= \hat{P}_t - \sum_{i=1}^{t-1} S(t - i + 1) Z_i.
\label{inc-prev:eq:forward-substitute}
\end{align}

>>>>

Considering multiple values of $t$ makes \cref{inc-prev:eq:EPt} into a system of simultaneous equations.
Therefore, the system can be written in matrix form:
\begin{align}
  \E(\vec{P} \mid \vec{Z}) = \matr{S} \vec{Z}. \label{inc-prev:eq:EPt-matrix} \\
\intertext{$\matr{S}$ is a lower-triangular matrix formed from $S$:}
    (\matr{S})_{i,j} &= \begin{cases}
        S(i - j + 1) & i - \dmax - 1 \leq j \leq i \\
        0 & \text{otherwise.}
    \end{cases} \label{inc-prev:eq:Smatrix}
\end{align}
In \cref{E-backcalc}, I will need to solve for this system for $\vec{Z}$ given a known $\hat{P} = \E(\vec{P} \mid {Z})$.
Assuming $\vec{P}$ and $\vec{Z}$ have the same dimension then $\matr{S}$ is lower-triangular with strictly positive diagonal elements; therefore, $\matr{S}$ is invertible, and the system has a unique solution~\autocite[424]{bernsteinFacts}.
The lower-triangular structure of $\matr{S}$ makes allows solving the system through forward-substitution, an efficient method that avoids explicit matrix inversion~\autocite{cormenMatrix}.
Forward-substitution sequentially solves the $t$th element of $\vec{Z}$ ($t = 1, \dots, T$) by calculating:
\begin{align}
Z_t
&= \frac{\hat{P}_t - \sum_{i=1}^{t-1} S_{ti} Z_i}{S_{tt}} \\
&= \hat{P}_t - \sum_{i=1}^{t-1} S(t - i + 1) Z_i.
\label{inc-prev:eq:forward-substitute}
\end{align}

### 2024-03-27 transmission-appendix-INLA.tex
finish proof reading

For INLA to perform well, \ie be compuutationally efficient and an accurate approximation, $\vec{\theta}$ should be of low dimension (\textcite{martinoINLAtutorial} recommend $< 15$); and the precision matrix, $\matr{\Sigma}^{-1}$, should be sparse.
However, the dimension of $\vec{x}$ is unrestricted, allowing a large number of random effects, or more complex structures.
The low dimensionality of $\vec{\theta}$ is because the method requires numerical integration over it.
The sparsity of $\matr{\Sigma}^{-1}$ allows for efficient computation.
Sparsity is introduced by conditional independence structures of the latent field that are present in many models.
For example, in the classic random effect setting, all random effects are independent conditional on the hyperparameters.

@@

Various methods exist to sample from the joint posterior of the hyperparameters and the latent variables, based on the INLA approximation.
The joint posterior is not approximated by default.
I used the one implemented in the R-INLA function \texttt{inla.posterior.sample}, described in \textcite[section 4]{chiuchioloJoint}.

>>>>

For INLA to perform well, \ie be compuutationally efficient and an accurate approximation, $\vec{\theta}$ should be of low dimension (\textcite{martinoINLAtutorial} recommend $< 15$); and the precision matrix, $\matr{\Sigma}^{-1}$, should be sparse.
However, the dimension of $\vec{x}$ is unrestricted, allowing a large number of random effects, or more complex structures.
The low dimensionality of $\vec{\theta}$ is because the method requires numerical integration over it.
The sparsity of $\matr{\Sigma}^{-1}$ allows for efficient computation.
Sparsity is introduced by conditional independence structures of the latent field that are present in many models.
For example, in the second-order random walk from \cref{biology-data:sec:MRP}, $w_t$ is independent from the rest of the parameters conditional on $w_{t-1}$, $w_{t-2}$, $w_{t+1}$, $w_{t+2}$.

@@

Various methods exist to sample from the joint posterior of the hyperparameters and the latent variables, based on the INLA approximation.
The joint posterior is not approximated by default.
I use the implementation in the R-INLA function \texttt{inla.posterior.sample}, described in \textcite[section 4]{chiuchioloJoint}.

### 2024-03-27 transmission-appendix-phase-type.tex
finish proof reading

Deciding $l$ and the length of time each individual spends in each compartment is the most challenging part of formulating the observation model.
The class of distributions which can be represented as a series of compartments are known as \emph{phase-type} distributions.
Discussions of general phase-type distributions are beyond the scope of this thesis; a review in a biological context can be found in \textcite{hobolthPhasetype}.
I will restrict myself to the subset of phase-type distributions defined by \textcite{osogamiClosed} as Erlang-Coxian (EC) distributions.
EC distributions are convenient because they are flexible enough to approximate almost any distribution, and the parameters of the approximation are given by closed-form functions of the distribution's moments.

@@

A total of five parameters are required to specify a $l$-phase EC distribution.
\begin{itemize}
    \item $l$, the number of phases.
    \item $\omega_Y$, the rate parameter of the Erlang distribution.
    \item $\omega_{X1}$, the rate parameter of the first state after the Erlang distribution.
    \item $\omega_{X2}$, the rate parameter of the second state after the Erlang distribution.
    \item $\pi$, the probability of moving from the end state of the Erlang distribution to the second (as opposed to directly to the absorbing state).
\end{itemize}

>>>>

Deciding $l$ and the length of time each individual spends in each compartment is the most challenging part of formulating the observation model in \cref{E-SEIR:sec:observation}.
The class of distributions which can be represented as a series of compartments are known as \emph{phase-type} distributions.
Discussions of general phase-type distributions are beyond the scope of this thesis; a review in a biological context can be found in \textcite{hobolthPhasetype}.
I will restrict myself to the subset of phase-type distributions defined by \textcite{osogamiClosed} as Erlang-Coxian (EC) distributions.
EC distributions are convenient because they are flexible enough to approximate almost any distribution, and the parameters of the approximation are given by closed-form functions of the distribution's moments.

@@
